A fast and memory-efficient transition state search method for complex energy landscapes

### 2022-04-18 manuscript.tex
Add discussion of hybrid methods to conclusion, modify string/DNEB, add distance metric.

The method begins by first initialising the states, $\bm{x}_1$ and $\bm{x}_2$, in the basins of attraction of different local minima, such as the two blue spots in the 2d potential in \cref{fig:toy2d}a.
These can be set to the minima, but this is not a necessary requirement.
The energies of these two states are then minimised, while constraining their separation.
This is iteratively reduced to zero, such that, at iteration $i$, their separation is
\begin{equation}\label{eq:diteration}
  d_i = (1 - f) d_{i-1}.
\end{equation}
With $d_0$ as the separation between the two initial states.
A reduction factor of $f = 0.5$ is successful for most applications, but this can be made smaller to ensure that the states do not slide off the ridge between the two basins of attraction.
Different metrics may be used to compute the distance.
%######
To further ensure that neither state is pulled over the ridge, a secondary constraint enforces equal energies for the two states.
Using this strategy, the two states will meet at the lowest point on the ridge - the transition state.

The two constraints are implemented using energy penalty terms, which result in driving forces on the two states if the constraints are not met, such as in \cref{fig:toy2d}b.
Including these energy penalty terms gives the total BITSS energy for the pair of states,
\begin{multline}\label{eq:bitss}
  E_\text{BITSS}(\bm{x}_1, \bm{x}_2) = E_1 + E_2
    + \kappa_e \left( E_1 - E_2 \right) ^2 \\
    + \kappa_d \left( \mathrm{d}(\bm{x}_1, \bm{x}_2) - d_i \right) ^2,
\end{multline}
where $E_1$ and $E_2$ are the single state energies, $\kappa_e$ and $\kappa_d$ parametrise the strengths of the energy and distance constraints, and $\mathrm{d}(\bm{x}_1, \bm{x}_2)$ is the measure of distance between the two states.

@@

Once BITSS has identified the transition state, it is possible to find the full minimum energy pathway by tracing the trajectory of downhill minimisations from the two final states, either side of the transition state.
In the case where there is a chain of transition states between the two starting minima, BITSS will identify the transition state with the highest energy in the chain.
If all of transition states or the full pathway are desired instead, BITSS can be continually repeated from one of the minima downhill from the located transition state and one of the initial minima until the initial minima are piecewise connected by a full pathway.

@@

\section{Conclusion}
Overall, we have developed the binary image transition state search (BITSS) algorithm for the efficient location of transition states in traditionally challenging landscapes.
Principally, the demonstrated speed and memory-efficiency will be key as we move towards simulating systems with increasingly large numbers of degrees of freedom, such as in large biomolecules, or with increasingly fine resolution, such as in finite element domains.

A second source of efficiency in the BITSS method comes from the ability to adaptively change the degrees of freedom as the algorithm proceeds.
We demonstrated how transition states could be found by both increasing the resolution upon convergence, and coupling systems with different discretisations.
The ease of coupling two copies of a system and adaptive remeshing, now leads to the possibility of incorporating BITSS into existing optimisation methods, such as surface evolver or finite element analysis software, to provide important energy barrier functionality.

Finally, we showed how BITSS can be used to survey discontinuous energy landscapes, demonstrated for a system of attractive hard-core particles.
This opens up possibilities for studying a broad range of systems previously out of reach of conventional landscape methods, but where transition information is valuable.
These include systems with very short range interactions or contact forces, such as in colloidal clusters, wrinkling and buckling of elastic materials, or locomotion and environmental interaction in robotics.

The distance metric between the two BITSS images proves interesting to analyse further in the future.
One question that emerges is whether transition states can be located by coupling two images through a small number of collective properties, rather than the total distance between all degrees of freedom in the system.
A second question is that for landscapes with multiple pathways between states, how can the distance metric be biased to access transition states different from the most direct one? Overall, these will enable BITSS to access even more challenging landscapes, and those not yet amenable to traditional landscape exploration techniques.

@@

@@

%% start bib
% \bibliography{bibexport.bib}
% % \bibliography{../../library.bib}
% % \bibliographystyle{apsrev4}
%% mid bib
\begin{thebibliography}{50}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
\expandafter\ifx\csname bibnamefont\endcsname\relax
  \def\bibnamefont#1{#1}\fi
\expandafter\ifx\csname bibfnamefont\endcsname\relax
  \def\bibfnamefont#1{#1}\fi
\expandafter\ifx\csname citenamefont\endcsname\relax
  \def\citenamefont#1{#1}\fi
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
\providecommand{\bibinfo}[2]{#2}
\providecommand{\eprint}[2][]{\url{#2}}

@@

>>>>

The method begins by first initialising the states, $\bm{x}_1$ and $\bm{x}_2$, in the basins of attraction of different local minima, such as the two blue spots in the 2d potential in \cref{fig:toy2d}a.
These can be set to the minima, but this is not a necessary requirement.
The energies of these two states are then minimised, while constraining their separation.
This is iteratively reduced to zero, such that, at iteration $i$, their separation is
\begin{equation}\label{eq:diteration}
  d_i = (1 - f) d_{i-1}.
\end{equation}
With $d_0$ as the separation between the two initial states.
A reduction factor of $f = 0.5$ is successful for most applications, but this can be made smaller to ensure that the states do not slide off the ridge between the two basins of attraction.
Different metrics may be used to compute this distance, although in this work we simply use the Euclidean distance,
\begin{equation}
  \mathrm{d}(\bm{x}_1, \bm{x}_2) = \sqrt{\sum_i (x_{1,i} - x_{2,i})^2}.
\end{equation}
To further ensure that neither state is pulled over the ridge, a secondary constraint enforces equal energies for the two states.
Using this strategy, the two states will meet at the lowest point on the ridge - the transition state.

The two constraints are implemented using energy penalty terms, which result in driving forces on the two states if the constraints are not met, such as in \cref{fig:toy2d}b.
Including these energy penalty terms gives the total BITSS energy for the pair of states,
\begin{multline}\label{eq:bitss}
  E_\text{BITSS}(\bm{x}_1, \bm{x}_2) = E_1 + E_2
    + \kappa_e \left( E_1 - E_2 \right) ^2 \\
    + \kappa_d \left( \mathrm{d}(\bm{x}_1, \bm{x}_2) - d_i \right) ^2,
\end{multline}
where $E_1$ and $E_2$ are the single state energies, and $\kappa_e$ and $\kappa_d$ parametrise the strengths of the energy and distance constraints.

@@

Once BITSS has identified the transition state, it is possible to find the full minimum energy pathway by tracing the trajectory of downhill minimisations from the two final states, either side of the transition state.
In the case where there are intermediaries, resulting in a chain of multiple transition states between the two minima, BITSS will identify the transition state with the highest energy in the chain.
This enables the identification of the overall energy barrier, providing estimates about the overall ease of the transition, or the rate for chemical processes.
However, if all of transition states or the full pathway are desired, BITSS can be continually repeated from one of the minima downhill from the located transition state and one of the initial minima until the initial minima are piecewise connected by a full pathway.

@@

\section{Conclusion}
Overall, we have developed the binary image transition state search (BITSS) algorithm for the efficient location of transition states in traditionally challenging landscapes.
This has distinct advantages for complex pathways owing to the lack of a required initial pathway estimate, as well as the identification of the transition state that provides the overall energy barrier in multi-step pathways.
Furthermore, the demonstrated speed and memory-efficiency will be key as we move towards simulating systems with increasingly large numbers of degrees of freedom, such as in large biomolecules, or with increasingly fine resolution, such as in finite element domains.

A second source of efficiency in the BITSS method comes from the ability to adaptively change the degrees of freedom as the algorithm proceeds.
We demonstrated how transition states could be found by both increasing the resolution upon convergence, and coupling systems with different discretisations.
The ease of coupling two copies of a system and adaptive remeshing, now leads to the possibility of incorporating BITSS into existing optimisation methods, such as surface evolver \cite{Brakke1992} or finite element analysis software, to provide important energy barrier functionality.

Finally, we showed how BITSS can be used to survey discontinuous energy landscapes, demonstrated for a system of attractive hard-core particles.
This opens up possibilities for studying a broad range of systems previously out of reach of conventional landscape methods, but where transition information is valuable.
These include systems with very short range interactions or contact forces, such as in colloidal clusters, wrinkling and folding of elastic materials, or locomotion and environmental interaction in robotics.

During the comparison with other methods, we have not considered hybrid methods, which involve switching to a single-ended method when close to the transition state, for greater accuracy and efficiency.
While BITSS does not need to use such a hybrid method to accurately find the transition state, it could be utilised to further increase the efficiency of the algorithm.
Therefore, future work is to look into how best to implement this and see how it compares against other hybrid methods.

The distance metric between the two BITSS images also proves interesting to analyse further.
One question that emerges is whether transition states can be located by coupling two images through a small number of collective properties, rather than the total distance between all degrees of freedom in the system.
A second question is that for landscapes with multiple pathways between states, how can the distance metric be biased to access transition states different from the most direct one?
Overall, these will enable BITSS to access even more challenging landscapes, and those not yet amenable to traditional landscape exploration techniques.

@@

\subsection{Implementing the string and DNEB methods}
In the convergence rate comparison in \cref{fig:speedtest}, the string method and doubly-nudged elastic band method (DNEB) are used, based upon \cite{E2007} and \cite{Trygubenko2004}, respectively.
Each method is initialised with 5, 10, or 20 images, using a linear interpolation between the chosen end states.

In each iteration of the string method, each image is evolved downhill for a number of LBFGS iterations before redistributing the images along a cubic-spline interpolated path.
This number of iterations is varied to obtain the fastest convergence, for which we find 2, 50, and 20 to be optimal for the particle cluster, cylidrical buckling, and striped wetting examples, respectively.

In the case of DNEB, the spring constant between the images is instead optimised.
For the particle cluster system, a spring constant of $1$ is used for the 5 image case, and $0.1$ for 10 and 20 images.
For the cylindrical buckling, the spring constant is $0.01$ in all cases.
And for the striped wetting, $10^{-4}$ is used for 5 and 10 images, and $10^{-3}$ for 20 images.

@@

%% start bib
% \bibliography{bibexport.bib}
% % \bibliography{../../library.bib}
% % \bibliographystyle{apsrev4}
%% mid bib
\begin{thebibliography}{51}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
\expandafter\ifx\csname bibnamefont\endcsname\relax
  \def\bibnamefont#1{#1}\fi
\expandafter\ifx\csname bibfnamefont\endcsname\relax
  \def\bibfnamefont#1{#1}\fi
\expandafter\ifx\csname citenamefont\endcsname\relax
  \def\citenamefont#1{#1}\fi
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
\providecommand{\bibinfo}[2]{#2}
\providecommand{\eprint}[2][]{\url{#2}}

@@

\bibitem[{\citenamefont{Brakke}(1992)}]{Brakke1992}
\bibinfo{author}{\bibfnamefont{K.~A.} \bibnamefont{Brakke}},
  \bibinfo{journal}{Experimental Mathematics} \textbf{\bibinfo{volume}{1}},
  \bibinfo{pages}{141} (\bibinfo{year}{1992}).
